---
title: "Impact of chess elo ratings"
---


LOREM IPSUM INTRODUCTIO ELOREM

#### Hypothesis
1. Games with higher rating more likely end with resign than games with lower rating
2. At lower ratings the difference between black and white is neglegible
3. The higher the rating the more likely games are to have more turns.
4. Games with lower time control end more likely with mate or out of time.

LOREM IPSUM

Approach 1:
What stats do we want:
- winner_rating
- loser_rating
Plot winner_rating - loser_ratings over winner_rating / loser_rating

What are we doing with draws?

Approach 2:
We need 
- underdog_rating
- topdog_rating
- winner_dog
- diff = topdog_rating - underdog_rating
Plot histogram of loses draws and wins.




Apprach Kaffepause:

1. Plot Histogram (topdog_rating - underdog_rating) 


#### Evaluation
##### Prerequisits
``` {r message=FALSE}
library(remotes)
options(repos = "https://cloud.r-project.org/")
if (!require("easypackages")) {
  install.packages("easypackages")
  library(easypackages)
} else {
  library(easypackages)
}
packages(
  "tidyverse",
  "tidytuesdayR",
  "dplyr",
  "readr",
  prompt = FALSE
)
```
##### Loading data
```{r message=FALSE}
tuesday_chess <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-01/chess.csv')
head(tuesday_chess)
```
##### Preprocessing
```{r}
chess <- tuesday_chess %>%
  dplyr::filter(rated=TRUE) %>%
  dplyr::select(winner, white_rating, black_rating, white_id, black_id, end_time)
head(chess)

extra <- chess %>% 
  dplyr::filter(winner != "draw") %>%
  dplyr::mutate(winner_rating = ifelse(winner == "white", white_rating, black_rating)) %>%
  dplyr::mutate(loser_rating = ifelse(winner == "white", black_rating, white_rating))
  
extra <- extra %>%
  mutate(
    rating_diff = winner_rating - loser_rating,  # Difference between winner and loser rating
    rating_ratio = loser_rating / winner_rating  # Ratio of winner to loser rating
  )


# Create the plot
ggplot(extra, aes(x = rating_ratio, y = rating_diff)) +
  geom_point(alpha = 0.6, color = "steelblue") +  # Scatter plot
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Optional trendline
  labs(
    title = "Rating Difference vs. Rating Ratio",
    x = "Winner Rating / Loser Rating",
    y = "Winner Rating - Loser Rating"
  ) +
  theme_minimal()



player_ratings <- chess %>%
  pivot_longer(
    cols = c(white_rating, black_rating),
    names_to = "pivot_label",
    values_to = "rating"
  ) %>%
  mutate(player_id = ifelse(pivot_label == "white_rating", white_id, black_id)) %>%
  select(player_id, rating, end_time) %>%
  group_by(player_id) %>%
  summarise(
    max_rating = max(rating),
    min_rating = min(rating),
    avg_rating = mean(rating),
    rating = rating[which.max(end_time)],  
    games_played = n()
  ) %>%
  ungroup()
head(player_ratings)

ggplot(player_ratings, aes(x = rating)) +
  geom_histogram(
    aes(y = after_stat(count) / sum(after_stat(count)) * 100),
    binwidth = 50,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram of Average Player Ratings",
    x = "Rating",
    y = "Percentage of Players"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal()



```




#### Conclusion
LOREM ISPUM
