---
title: "Impact of chess elo ratings"
bibliography: references.bib
csl: apa.csl
---
# Abstract
TODO: AT THE END!! :-)

=^)

[@rating]

[@1]

# Preface
In this paper, we present a brief empirical analysis of the dataset from Week 40 of the Tidytuesday project. Tidytuesday is an open-source initiative within the R community that encourages participants to conduct small-scale analyses and create visualizations based on datasets provided weekly.

The Week 40 dataset originates from a popular online chess platform and includes a subset of games played on the site.

This work was completed as part of the 3-day seminar "Data Literacy in R" at KIT. As the contributors are neither experts in data science nor chess, we advise against using or referencing this paper in academic or professional contexts.

# Introduction
## First move advantage
The concept of the first-move advantage in chess has been widely recognized since 1889, when the first World Chess Champion, Wilhelm Steinitz, expressed concerns that further analysis of the game might reveal an inherent advantage for White, potentially leading to a "solved" and stagnant game.

Empirical evidence from subsequent World Chess Championships supports the existence of this advantage. Historically, White's win rate has been consistently higher than Black's by approximately 3 to 5 percentage points. This dynamic has influenced competitive strategy, with Black often aiming to neutralize White's initiative and secure a draw. For example, the Berlin Defense has gained notoriety for forcing White to either relinquish its opening advantage or enter draw-prone positions.

Modern chess engines, such as Stockfish, further illustrate the complexities of the first-move advantage. When Stockfish plays against itself, games where White wins are more frequent than games where Black wins, although the vast majority of games still end in a draw. This pattern highlights White's ability to retain a slight edge in optimal play scenarios, even against a perfectly rational opponent.

At the highest levels of chess, this phenomenon has contributed to concerns about the game's stagnation. Bobby Fischer, one of the greatest chess players in history, introduced Fischer Random Chess (also known as Chess960) as an alternative. By randomizing the initial piece setup, Fischer sought to reduce reliance on opening theory and emphasize creativity and tactical skill. Notably, the first-move advantage in Fischer Random is less pronounced, as even professional players tend to make less optimal moves in uncharted positions, diminishing White's edge.

Today, the first-move advantage is universally acknowledged by players and theorists alike. Competitive preparation often involves strategies and openings specifically designed to address this imbalance. Moreover, it is theorized—and supported by empirical data—that the first-move advantage becomes more significant as players approach near-perfect play.


## Chess Rating system
The original chess rating system, known as the Elo system, was introduced in 1950 by Arpad Elo. It assigns players a rating typically ranging from 1000 to 2800 for most competitive players, though theoretically, ratings can extend beyond this range. The system models a player’s skill as a single numerical value and predicts game outcomes based on the difference between two players' ratings. Specifically, it assumes a Gaussian (normal) distribution of player skill, with the probability of victory determined by a logistic function of the rating difference. The Elo system remains in use today by FIDE (the International Chess Federation) for over-the-board chess tournaments.

While the Elo system is widely used, advancements in data collection and research on predictive models have led to the development of more sophisticated rating systems. One such system is Glicko, introduced by Mark Glickman in 1995. Glicko is currently employed by chess.com, one of the largest online chess platforms. Though originally developed for chess, Glicko has been adapted for use in ranking players in other games, such as Counter-Strike, Dota 2, and Team Fortress 2.

An enhanced version of Glicko, known as Glicko-2, was later developed to include a "rating deviation" (RD) component, which accounts for the confidence level of a player's rating. To our knowledge, the only major chess platform currently using Glicko-2 is Lichess.

Because these systems differ in implementation and starting points, comparing ratings between platforms can be misleading. For instance, Glicko-1 typically sets new players at an initial rating of 1000, representing the average skill level, while Glicko-2 starts players at 1500, reflecting a different baseline. As a result, direct comparisons between ratings from different systems are not meaningful.

## Over-the-board chess etiquet
In professional over-the-board chess, an interesting aspect of etiquette is the expectation to resign in clearly losing positions. Continuing to play in such situations can be perceived as disrespectful, as it implies doubt about the opponent's ability to convert an obvious advantage into victory. This expectation stems from the assumption that both players possess the skill and knowledge to recognize decisive positions, and prolonging the game unnecessarily might be seen as wasting time or challenging the opponent's competence.

This cultural norm contrasts with casual or beginner-level play, where players are often encouraged to play to checkmate, as their opponent's ability to find a winning sequence may not be guaranteed.

## Time Control
TODO
LOREM IPSUM OLOREM

## Hypothese

# Preliminary Analysis
In this section we want to provide a short look at the lichess dataset to 
evaluate its completeness. 
In the previous section about the first move advantage, we found that in 
professional over the board games it is generally an accepted truth that 
the player with the white pieces has a advantage, we now aim to verify that this 
assumption is also true for our lichess dataset. 
To evaluate this assumption we simly compute the average win percentage across 
all rated games in our dataset.
```{r message=FALSE, echo=FALSE, fig.align='center'}
rm(list = ls())
library(remotes)
options(repos = "https://cloud.r-project.org/")
options(scipen = 999)
if (!require("easypackages")) {
  install.packages("easypackages")
  library(easypackages)
} else {
  library(easypackages)
}
packages(
  "tidyverse",
  "tidytuesdayR",
  "dplyr",
  "readr",
  prompt = FALSE
)

tuesday_chess <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-01/chess.csv')

# Filter and select relevant columns
chess <- tuesday_chess %>%
  filter(rated == TRUE) %>% # Fix logical comparison
  select(winner)

# Converts the chess table of column winner into a 3 columns, which count the
# amount of times a game ended with white winning, black winning or
# the game beeing drawn
outcomes <- table(chess$ winner)

outcome_percentages <- prop.table(outcomes) * 100

pie(outcome_percentages,
    labels = paste(
      names(outcome_percentages),
      "(", round(outcome_percentages, 1), "%)", sep = ""),
    col = c("lightblue", "lightgreen", "lightcoral"),
    main = "Lichess dataset rated games outcome")
```

*The full source code is available [here](source.html)*

This short preliminary analysis shows that the first move advantage is clearly
present in our small lichess data set, which might indicate that the subsection of 
overall games played on lichess represented in the data set if not screwed 
in this regard.


Next up in our background section we found that lichess is using the Glicko II system, 
therefor our rating distribution across games should represent a normal distribution 
around the average rating of 1500, we can verify this assumption by 
looking at the ratings of all players in our data set.
For this we have to be careful to not count ratings of players with more games,
twice as those are likely to be rated higher. Luckily our data set contains a 
player id which we can use to only count the ratings of players once.
If multiple games for a single player exists we can pick either the most recent rating 
or the average rating across games, as it is unclear to us if averaging the 
ratings might lead to a statitical error we choose to select the rating of the most 
recently played game.
``` {r message=FALSE, echo=FALSE, fig.align='center'}
rm(list = ls())

library(remotes)
options(repos = "https://cloud.r-project.org/")
options(scipen = 999)
if (!require("easypackages")) {
  install.packages("easypackages")
  library(easypackages)
} else {
  library(easypackages)
}
packages(
  "tidyverse",
  "tidytuesdayR",
  "dplyr",
  "readr",
  prompt = FALSE
)

tuesday_chess <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-01/chess.csv')

# Filter and select relevant columns
chess <- tuesday_chess %>%
  filter(rated == TRUE) %>% # Fix logical comparison
  select(white_id, white_rating, black_id, black_rating, start_time)

player_ratings <- chess %>%
  pivot_longer(
    cols = c(white_rating, black_rating),
    names_to = "pivot_label",
    values_to = "rating"
  ) %>%
  mutate(player_id = ifelse(pivot_label == "white_rating", white_id, black_id)) %>%
  select(player_id, rating, start_time) %>%
  group_by(player_id) %>%
  summarise(
    max_rating = max(rating),
    min_rating = min(rating),
    avg_rating = mean(rating),
    rating = rating[which.max(start_time)],
    games_played = n()
  ) %>%
  ungroup()

ggplot(player_ratings, aes(x = rating)) +
  geom_histogram(
    aes(y = after_stat(count) / sum(after_stat(count)) * 100),
    binwidth = 100,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Average Player Ratings",
    x = "Rating",
    y = "Percentage of Players"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal()
```

*The full source code is available [here](source.html)*

This short preliminary analysis, is yet another indicator to the completeness 
of the lichess data set, indicating that the games represented in the dataset 
do not represent any screw towards higher or lower rated games.

**TODO section about time control distribution in the dataset**

# Hypothesis
In this small dissertation, we aim to prove 2 hypothesis.

- **The first move advantage is more prominant in higher rated games.**
- **Higher rated players are more likely to resign than to be checkmated.**

### Influence of rating on the first move advantage
As seen in our preliminary research, we found it to be likely that,
the first move advantage becomes more significant as the level of 
play increases, this might be indicated by gamemodes like Chess960,
where the first move advantage is less significant. 
Further we additionally found sources that state that the first move
advantage is less likely to be prominent in Blitz games, which might
also indicate that the level of play as a direct influence on the 
strength of the first move advantage. 
Another interessting component of this hypothesis is that compared to our
preliminary research this work focuses on online chess statistics compared 
to over the board, where players regardless of their rating might be less
concentrated, further as lichess compared to other online platforms like 
chess.com does not play a significant role in the chess community 
players might treat their rating more casually, which might lead to 
less optimal play even at higher ratings.

### Influence of rating on game outcomes.
TODO

# Methods
For the evaluation we used RStudio with the tidyverse packaged....

# Results
Stumpf vorlesen

# Discussion
TODO

# Conclusion
TODO


# Contributors
- Karl Sassie

# Literatur
